Do researchers preferentially collaborate with same-gendered colleagues?

### 2018-05-01 manuscript/Manuscript.Rmd
Intro

# Introduction
Women are substantially underrepresented in the workforce in many branches of science, technology, engineering, mathematics, and medicine (STEMM), and they face additional challenges and inequities relative to men [e.g. @Lariviere_2013; @West_2013; @Elsevier_report; @Holman_2018]. On average, women occupy more junior positions (Wutte 2007, Reuben, Sapienza & Zingales, 2014) with lower salaries (Trower and Chait 2002, Umback 2007), receive less grant money (Hosek et al. 2005, OER 2005), are promoted more slowly (Hopkins et al. 2013; Long et al. 1993, Rosenfeld 1991, Zuckerman 1987), and are allocated fewer resources (European Commission 2009) and less research funding (Feldt 1986, Stack 2004, Larivière et al. 2011) than men. Experimental studies have demonstrated that researchers regard women's achievements less favourably than identical achievements by men (Moss et al. 2012; Knobloch et al. 2013). Multiple studies have found that women publish fewer papers per year than men (Cole and Zuckerman 1984, Long 1992, Larivière et al. 2013, Astronomy one in Nature) [check whether they control for career stage].

Because publishing, networking and collaboration are all instrumental to academic career advancement (Lee and Boheman 2005, Hopkins et al. 2012), numerous studies have tested for gender differences in these areas (reviewed in Table S1 in Holman et al. 2018; see also XXX). For example, studies have shown that women tend to be less involved in international collaboration than men (Lemoine 1995; Lewison 2001; Webster 2001; Larivière et al. 2011, Bozeman and Corley 2004), collaborate less with researchers from their own university department (Websters 2001), and have less prestigious collaborations compared to men (Long 1990), but also fewer collaborations in general (Fuchs Von Stebut and Allmendinger 2001). Possible explanations for the smaller average professional networks of women and include a greater average amount of family obligations (Long 1990, Wright et al. 2003, Reskin 1978), lower participation of women in international conferences (Lewison 2001, Eisen 2014, Martin 2014), and reduced receipt of research travel funds by women (Bozeman and Corley 2004).

@@

However, we believe that all of these studies used sampling strategies that could overstate the strength of homophily, making the true extent of homophily difficult to ascertain from present data. Essentially, whenever samples are taken from two or more discrete sets of literature that vary in the author gender ratio and that are largely not connected by collaboration, the number of same-gendered coauthors will be inflated (\autoref{wahlund_plot}). This can make it appear as though authors preferentially publish with same-gendered colleagues if no gender preferences exist, or even if researchers prefer opposite-gendered colleagues. For example, a sample of bioinformatics and cell biology papers might contain an excess of mostly-male and mostly-female author lists, simply because researchers preferentially work with colleagues from the same discipline, and because more women work in cell biology than in bioinformatics [@Holman_2018]. We name this issue the Wahlund effect, by analogy with the Wahlund effect in population genetics (Wahlund 1928), whereby spatial differences in allele frequencies create an excess of homozygotes across the whole population, even if alleles assort randomly within each subpopulation.

>>>>

# Introduction
Women are substantially underrepresented in many branches of the workforce in science, technology, engineering, mathematics, and medicine (STEMM), and they face additional challenges and inequities relative to men [e.g. @Lariviere_2013; @West_2013; @Elsevier_report; @Holman_2018]. On average, women occupy more junior positions (Wutte 2007, Reuben, Sapienza & Zingales, 2014) with lower salaries (Trower and Chait 2002, Umback 2007), receive less grant money (Hosek et al. 2005, OER 2005), are promoted more slowly (Hopkins et al. 2013; Long et al. 1993, Rosenfeld 1991, Zuckerman 1987), and are allocated fewer resources (European Commission 2009) and less research funding (Feldt 1986, Stack 2004, Larivière et al. 2011) than men. Experimental studies have demonstrated that researchers regard women's achievements less favourably than identical achievements by men (Moss et al. 2012; Knobloch et al. 2013).

Because publishing, networking and collaboration are all instrumental to academic career advancement (Lee and Boheman 2005, Hopkins et al. 2012), numerous studies have tested for gender differences in these areas (reviewed in Table S1 in Holman et al. 2018; see also XXX). For example, studies have shown that relative to men, women tend to be less involved in international collaboration (Lemoine 1995; Lewison 2001; Webster 2001; Larivière et al. 2011, Bozeman and Corley 2004), collaborate less with researchers from their own university department (Websters 2001), have less prestigious collaborations (Long 1990), and fewer collaborations in general (Fuchs Von Stebut and Allmendinger 2001). Possible explanations for the smaller average professional networks of women include a greater average amount of family obligations (Long 1990, Wright et al. 2003, Reskin 1978), lower participation of women in international conferences (Lewison 2001, Eisen 2014, Martin 2014), and reduced receipt of research travel funds by women (Bozeman and Corley 2004).

@@

However, we hypothesise that these studies of gender homophily were hindered by a sampling issue that we term the Wahlund effect (\autoref{wahlund_plot}), making the true extent of homophily difficult to ascertain from present data. Essentially, whenever samples are taken from two or more discrete sets of literature, which vary in the author gender ratio and which are largely not connected by collaboration, the number of same-gendered coauthors will be inflated. This can make it appear as though authors preferentially publish with same-gendered colleagues if no gender preferences exist, or even if researchers preferentially select opposite-gendered colleagues. For example, a sample of bioinformatics and cell biology papers might contain an excess of mostly-male and mostly-female author lists, simply because researchers preferentially work with colleagues from the same discipline, and because more women work in cell biology than in bioinformatics [@Holman_2018]. We name this issue after the Wahlund effect in population genetics (Wahlund 1928), whereby spatial differences in allele frequencies create an excess of homozygotes across the whole population, even if alleles assort randomly within each subpopulation.
